---
title: 'Case Control Study: Death on the Roads'
author: "Timothy Lee"
date: "11/04/2020"
output:
  pdf_document: default
  html_notebook: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Context
The dataset below is a subset of the data from `www.gov.uk/government/statistical-datasets/ras30-reported-casualties-in-road-accidents`, with all of the road traffic accidents in the UK from 1979 to 2015. The data below consist of all pedestrians involved in motor vehicle
accidents with either fatal or slight injuries (pedestrians with moderate injuries have been removed). We want to investigate whether the UK road accident data are consistent with the hypothesis that women tend to be, on average, safer as pedestrians than men, particularly as
teenagers and in early adulthood.

```{r, warning=FALSE, cache=TRUE}
pedestrainFile = Pmisc::downloadIfOld(
'http://pbrown.ca/teaching/303/data/pedestrians.rds')
pedestrians = readRDS(pedestrainFile)
pedestrians = pedestrians[!is.na(pedestrians$time), ]
pedestrians$y = pedestrians$Casualty_Severity == 'Fatal'
```

Dimensions of data
```{r}
dim(pedestrians)
```
Head of dataframe
```{r}
pedestrians[1:3, ]
```

Table of injuries by sex.
```{r}
table(pedestrians$Casualty_Severity, pedestrians$sex)
```

Range of time
```{r, cache=TRUE}
range(pedestrians$time)
```
## Preliminary Analysis
Notice that men are involved in accidents more than women, and the proportion of accidents which are fatal is higher for men than for women. This might be due in part to women being more reluctant than men to walk outdoors late at night or in poor weather, and could also
reflect men being on average more likely to engage in risky behaviour than women.

A `GLM` adjusting for weather and light conditions is below.
```{r, cache=TRUE}
theGlm = glm(y ~ sex + age + Light_Conditions + Weather_Conditions, data = pedestrians, family = binomial(link = "logit"))
knitr::kable(summary(theGlm)$coef, digits = 3)
```

Another `GLM` with interactions of sex and age is fitted. This model accounts for the significance of (the `interaction`) between both the person's `age` and `sex`.

```{r, cache=TRUE}
theGlmInt = glm(y ~ sex * age + Light_Conditions + Weather_Conditions,
data = pedestrians, family = binomial(link = "logit"))
knitr::kable(summary(theGlmInt)$coef, digits = 3)
```

## Plot of tredicted probability of being a case in baseline conditions (daylight, fine no wind) with 99% CI using `theGlmInt`.
```{r}
newData = expand.grid(
age = levels(pedestrians$age),
sex = c('Male', 'Female'),
Light_Conditions = levels(pedestrians$Light_Conditions)[1],
Weather_Conditions = levels(pedestrians$Weather_Conditions)[1])
thePred = as.matrix(as.data.frame(
predict(theGlmInt, newData, se.fit=TRUE)[1:2])) %*% Pmisc::ciMat(0.99)
thePred = as.data.frame(thePred)
thePred$sex =newData$sex
thePred$age = as.numeric(gsub("[[:punct:]].*|[[:alpha:]]", "", newData$age))
toPlot2 = reshape2::melt(thePred, id.vars = c('age','sex'))
toPlot3 = reshape2::dcast(toPlot2, age ~ sex + variable)

matplot(toPlot3$age, exp(toPlot3[,-1]),
type='l', log='y', col=rep(c('black','red'), each=3),
lty=rep(c(1,2,2),2),
ylim = c(0.007, 0.11), xaxs='i',
xlab= 'age', ylab='prob')
legend('topleft', lty=1, col=c('black','red'), legend = c('male','female'), bty='n')
```

## Discussion

For $theGlm$ object, the "case" group is defined to be the pedestrians involved in motor vehicle accidents with fatal injuries and the "control" group is defined to be ones with slight injuries. The intercept represents 26-35 year-old, male pedestrians under day light condition with fine weather and no high winds who are involved in motor vehicle accidents with either fatal or slight injuries. The covariates are sex (categorical with levels Female and Male, male is the reference level), age ( categorical with levels 0to5, 6to10, ... etc and 26to35 is the reference level), Light Conditions (categorical with levels darkness or daylight with lights lit, unlit, no lighting or lighting unknown and the daylight coniditon is our reference), and weather conditions (categorical with levels of raining, snowing, fine, fog or mist, high winds, or no high winds, and fine weather and no high winds is our reference).

For $theGlmInt$ object, the "case" group is defined to be the pedestrians involved in motor vehicle accidents with fatal injuries and the "control" group is defined to be ones with slight injuries. The intercept represents 26-35 year-old, male pedestrians under day light condition with fine weather and no high winds who are involved in motor vehicle accidents with either fatal or slight injuries. The covariates are sex (categorical with levels Female and Male, male is the reference level), age ( categorical with levels 0to5, 6to10, ... etc and 26to35 is the reference level), Light Conditions (categorical with levels darkness or daylight with lights lit, unlit, no lighting or lighting unknown and the daylight coniditon is our reference), weather conditions (categorical with levels of raining, snowing, fine, fog or mist, high winds, or no high winds, and fine weather and no high winds is our reference), and also the interaction terms of sex and age.


## Analysis of Hypothesis
In order to address the question of interest, we should consider model 2 as this model accounts for the potential interaction between age and sex (i.e., rather than only considering a particular age group or a particular sex individually, it makes more sense to consider a given age along with that age group's sex since different sexes in the same age group might vary greatly). This approach is also reinforced by the fact that only 3 of the age and sex interaction groups have "1" in the C.I., indicating a potential to have no effect. 

Hence, based on the data and results from model 2, it is possible to conclude that our findings are consistent to the fact that women tend to be, on average, safer as pedestrians than men. This is also shown by model 2's point estimate of an odds ratio 0.58 when sex = female, which is lower compared to the implicit reference level of sex=male with the odds ratio of 1. The fact that this estimate's confidence interval doesn't contain a "1", it is possible to conclude that there is a significant effect of sex (of females) to be less likely to be involved in an accident, making them "safer" compared to men.

However, this is not true/inconclusive for particularly women as teenagers and in early adulthood. If we define teenagers to early adulthood by the age of 16 to 35, we can see females in the first age group of 16 to 20 has a higher odds ratio of 1.16 (which is greater than "1" for the reference group, and "1" is also not in its CI, indicating that we are 95% confident that it has a significant effect). For the age group of 21 to 25, since "1" is in our confidence interval, we cannot be certain that this factor is significant. Fianlly, for the age group of 26 to 35 (which is our reference level), the odds ratio estimate is implicit to be 1. All in all, our findings do not support the fact that females are (significantly) safer particularly as teenagers and in early adulthood, but women do tend to be safer on average.

## Hypothesis 2: It is well established that women are generally more willing to seek medical attention for health problems than men, and it is hypothesized that men are less likely than women to report minor injuries caused by road accidents. We want to test that whether or not the control group is a valid one for assessing whether women are on average better at road safety than man.

Firstly, the hypothesis given is that men are less likely than women to report minor injuries caused by road accidents. Hence, based on figure 2, we can see that male does have a higher probability of being a case in baseline conditions (daylight, fine no winds) in $theGlmInt$ model. In other words, the data given supports the fact that men are more likely to be involved in accidents which are fatal, which could possibly reflect the fact that men are less likely than women to report minor injuries. However, there is a problem in the control group for assessing whether women are on average better at road safety than man, since some of the underlying assumptions of case-control studies could be violated such as the inclusion in the study might depend on the covariates of sex. For example, men and women might have different, subjective opinions on what is considered "fatal" and "slight" for injuries (due to perhaps physical reasons, where men might be more likely to describe more injuries as "slight" compared to females), which could in turn affect the results as our case control study could be based on the subject's reporting. Also, men might also be less likely to go directly to a hospital after either a fatal or slight injury (so these men won't be in our data in the first place), resulting in an under-representation of such men. Hence, due to these reasons of the control group, it might be inaccurate to generalize our findings of whether or not women are on average better at road safety than man.











